Small Language Modelling to Generate Playlists from Song Sequences

### 2023-12-11 docs/draft_report.tex
Updated Draft finalReport
Cleaned up Intro, Data Preparation sections

TODO: related work - recreation of original
TODO: clean up models sections
TODO: finish and integrate results discussion conclusion
TODO: Add visualizations and examples
TODO: Finish citations, footnotes

\title{Small Language Modelling for Song Sequence Generation}

@@

Phish, the iconic jam band hailing from Vermont since 1984, has captivated audiences with their improvisational rock and roll style, reminiscent of Jazz. Notably, their performances stand out for the absence of pre-planned setlists, creating a pseudo-random selection of songs during shows. This unique approach, while spontaneous, is subtly influenced by constraints like temporal frequency and positional sequencing.

In the pursuit of unraveling the intricate dynamics of Phish's musical repertoire, previous endeavors have tapped into a public API providing access to historical song and setlist data. These initiatives, showcased in works exploring network relationships between songs, have employed advanced techniques such as recurrent neural networks (RNN) with word2vec embeddings. Notably, these studies have achieved a validation accuracy of up to ~ 21\%, showcasing the potential of leveraging data-driven methodologies to predict the next song in a sequence.

This paper extends this trajectory by proposing a novel exploration into the world of artist-centric music recommendation systems, specifically tailored for dedicated fans. Drawing inspiration from Phish's improvisational ethos, our research endeavors to shed light on the compositional tendencies of the band. Through the application of cutting-edge neural network models, we aim to predict future setlists and song appearances, offering a fresh perspective on music streaming recommendations.

@@

@@

\section{Data Acquisition and Preparation}

For the acquisition of Phish concert data crucial to our predictive modeling endeavors, we adopt a methodical approach employing the Phish.net API. Armed with a unique API key provided by Phish.net, our initial focus centers on ingesting song-related details, encompassing song ID, artist, play frequency, last performance date, and debut date. Further, we meticulously filter shows attributed exclusively to Phish, forming a chronological compendium of their extensive concert history. The parsing of setlist data follows, extracting intricate details such as show date, set number, song position, ID, name, transition markers, gap since the last performance, and song categorization. Ensuring the integrity of our dataset, we eliminate incomplete information, resulting in a comprehensive Data Frame. This curated dataset, exclusively featuring "full" shows with two sets and an encore, serves as the foundational bedrock for our subsequent predictive modeling endeavors, poised to unveil patterns within Phish's musical evolution.

In the initial phase of data preprocessing for the Phish concert dataset, we conducted an insightful exploration to understand the diversity of songs. Statistical analyses were performed to calculate and present the distribution. To address the challenge posed by rarely performed songs, we consolidated songs with one or two performances into a unified category termed "wildcard." This involved reassigning identifiers (songid=0) and labels ('wildcard') for streamlined modeling.

Furthermore, for predictive modeling, we systematically aggregated songs into sequences by grouping them based on show date and set number. These sequences were then concatenated into a unified string labeled as "full" and aggregated by show date. This preprocessing step refines the dataset and establishes a structured framework for developing models focused on predicting Phish setlists.

\section{Procedures}

These sections review the scope of experiments performed as part of this work.

@@

Our data frame attribute “full” contains the songs from each concert. To prepare the data for tokenization the raw song data underwent a series of essential steps. Initially, the songs were tokenized, a process involving the division of the songs into individual words. This tokenization was accomplished by splitting the songs using the song delimiter ("|"), and additionally, as multi word songs had dashes, any occurrences of dashes were removed from the words. Subsequently, to establish a robust foundation for numerical representation, a vocabulary was constructed by compiling the unique set of words derived from all the songs. This vocabulary creation ensures that each word in the dataset possesses a distinct numerical identifier.  This indexing facilitated the creation of a mapping between words and their corresponding numerical representations, paving the way for numerical encoding.

This numerical conversion transforms the raw song data into a structured format suitable for input into a neural network model

Since our model has to predict either nth or next n songs, we employed multiple experimental splits and training models for each provides a comprehensive analysis of the model's performance under various conditions. The sequential split (Method 1) captures the immediate temporal dependencies within the songs. The time-based split (Method 2) considers the evolving nature of song sequences over time. The unified encoding approach (Method 3) offers insights into the global patterns and relationships across the entire song repertoire. This multi-faceted evaluation ensures a robust understanding of the model's predictive capabilities and generalizability across different scenarios.

\subsection{Models}

@@

\subsubsection{Unidirectional LSTM with N-Dimensional Embedding}

@@

\subsubsection{Bidirectional LSTM with N-grams and Adjacency Matrix Embedding}

@@

\subsubsection{Self-Attention Model with N-d + Positional Embedding}

@@

\subsubsection{Self-Attention + Cross-Attention Model with N-d + Positional Embedding}

@@

>>>>

\title{Small Language Modelling to Generate Playlists from Song Sequences}

@@

As of 2022, music streaming is an \$18 Billion dollar industry\footnote{\url{https://www.weforum.org/agenda/2023/03/charted-the-impact-of-streaming-on-the-music-industry/}}. Playlist generation is a key differentiator for streaming platforms and previous works have used reinforcement learning [1], convolutional neural networks (CNNs) [2], and recurrent neural networks (RNNs) [3] as the generative mechanisms. These works have generally spanned a broad scope of artists and genres to generate playlists composed of multiple bands or band-specific playlists ordered on an arbitrary learned feature. While this may work well for a generic music listener, this approach may not be effective for "dedicated fans" - a subset of users whose listening habits tend to be more focused around an individual band and their compositional tendencies.

One such type of “dedicated fan” are those followers of Jam bands. Since the 1960s, Jam bands have captivated audiences with their improvisational Rock and Roll style, reminiscent of Jazz. Notably, these groups do not pre-plan setlists for their live performances and instead weave together a pseudo-random selection of songs such that no two shows are alike. The choice of songs for each show is partially influenced by spontaneous notions such as crowd energy or a band member’s mood, but it is also influenced by constraints like temporal frequency and positional sequencing. These shows also generally follow a standard pattern composed of two sets separated by an intermission and a brief encore composed of just a song or two. Beyond the music itself, the sequencing of songs in shows is reminiscent of natural language – song names are analogous to words, setlists to sentences, show starts and stops to parts-of-speech – with the repertoire of each band forming its own unique language full of syntactical intricacies.

We purport that constructing small language models from the song sequences of live performances to generate playlists for a given band could increase the engagement of “dedicated fan” type users by generating playlists that more closely mimic a band’s musical tendencies than a more generalized algorithm. In this report, we explore the music of the Jam band Phish for this purpose – who has not only a large catalog of live shows and songs, but also a very dedicated fanbase. A previous endeavor\footnote{\url{https://towardsdatascience.com/predicting-what-song-phish-will-play-next-with-deep-learning-947ccce3824d}} employed a Long Short-Term Memory (LSTM) model with Word2Vec [4] algorithm for token embeddings to achieve a ~21\% accuracy for next song prediction given a sequence of Phish songs. This paper builds upon this work by recreating it four years later and exploring the prediction task with alternate embedding methods, and RNN model architectures, and attention model architectures.

@@

the work\footnote{\url{https://github.com/andrewrreed/phish-setlist-modeling}}

@@

\section{Data Preparation}

The \url{https://phish.net} fan forum curates an API\footnote{\url{https://docs.phish.net/}} cataloguing every setlist of songs for every show the band has played. First queried from this endpoint were song-related details encompassing song name, ID, artist, play frequency, last performance date, and debut date. Further, we meticulously filter shows attributed exclusively to Phish, forming a chronological compendium of their extensive concert history. Next queried were setlist data such as showdate, set number, song positions, ID, name, transition markers, song gaps since the last performance, and jam categorization. Incomplete information were eliminated to ensure the dataset's integrity, resulting in a comprehensive dataframe. This curated dataset, exclusively featuring "full" shows with two sets and an encore, serves as the foundation for the predictive modeling task and is saved as a csv file\footnote{\url{https://github.iu.edu/rjdharmc/dlProject/blob/main/data/allphishsets.csv}} in the project's code repository.

In the initial exploratory data analysis of the Phish concert dataset, it was revealed the frequency of songs played resembled a power-law distribution whereby almost half of the songs were only played once or twice throughout the band's entire history. To address the challenge posed by rarely performed songs, we consolidated these one-off and two-off songs into a unified song category termed the "wildcard". For prediction, this wildcard song is treated like any other song Phish plays but it represents an intractable uncertainty to relax the problem for new song debuts or unpredictably random cover songs.

Furthermore, for predictive modeling, we systematically aggregated songs into sequences by grouping them based on show date and set number. These sequences were then concatenated into a unified string with additional separators in the string to indicate a song's position in the show. For example, the beginning of every string starts with 'set-1'. Then after all the 1st set songs appear in the string, the 'set-2' separator appears, then the 'set-e' separator before the encore, and finally the 'eos' separator at the end of the string after all the songs.

These preprocessing steps refined the dataset and established a structured framework for developing models focused on predicting Phish setlists. Following this modification and the original data cleaning, the dataset is composed of 1,550 shows, 33,533 total songs played, with 482 unique songs and separators.

@@

In order to mathematically interpret the song names as a language, it was necessary to tokenize the song names and separators into numerical values suitable for input into a neural network model. This was predominantly done using unigram tokens, or integer representations mapped to each full song name. In this way, each setlist is similarly decomposed into a list of integer values.

Recall, the aim of this project is to generate song sequences (IE playlists) which are representative of the style in which the band plays. This problem can be framed in terms of language modelling in either a {\it sequence-to-token} perspective, using a sequence to predict the next token, or a {\it sequence-to-sequence} perspective, using a sequence to predict the next sequence. Since the models will have to either predict the next $n^{th}$ song or the next $n$ songs, multiple data splitting techniques for input and output data were applied to evaluate the model performance under both of these paradigms. These are described below.

\paragraph{Method 1}
The {\it sequential split} captures the immediate temporal dependencies within the songs. This approach involves a sequential splitting of the song data, where each setlist sequence is transformed into input and target sequences for training. The sequences are then padded to ensure uniform length. This method enables the model to learn and predict the next song in a sequence, essentially forecasting the playlist evolution

\paragraph{Method 2}
The {\it time-based split} considers the evolving nature of song sequences over time. In this approach, the dataset is chronologically split into training and testing sets based on a specified date, for example '2015-12-31'. This method reflects a time-based split, capturing the temporal dynamics of song sequences.

\paragraph{Method 3}
The {\it unified encoding approach} offers insights into the global patterns and relationships across the entire song repertoire. In this approach, the training data is like [n], [n, n+1], [n, n+1, n+2] and training labels are [n+1], [n+2], [n+3] and so on. Basically, the next song to the incoming sequence is our prediction label.

\section{Models}

@@

\subsection{Unidirectional LSTM with N-Dimensional Embedding}

@@

\subsection{Bidirectional LSTM with N-grams and Adjacency Matrix Embedding}

@@

\subsection{Self-Attention Model with N-d + Positional Embedding}

@@

\subsection{Self-Attention + Cross-Attention Model with N-d + Positional Embedding}

@@

	[1] F. Tomasi, J. Cauteruccio, S. Kanoria, K. Ciosek, M. Rinaldi, and Z. Dai, “Automatic Music Playlist Generation via Simulation-based Reinforcement Learning,” Proceedings of the 29th ACM SIGKDD Conference on Knowledge Discovery and Data Mining. ACM, Aug. 04, 2023. doi: 10.1145/3580305.3599777.

	[2] R. T. Irene, C. Borrelli, M. Zanoni, M. Buccoli, and A. Sarti, “Automatic playlist generation using Convolutional Neural Networks and Recurrent Neural Networks,” 2019 27th European Signal Processing Conference (EUSIPCO). IEEE, Sep. 2019. doi: 10.23919/eusipco.2019.8903002.

	[3] K. Choi, G. Fazekas, and M. Sandler, “Towards Playlist Generation Algorithms Using RNNs Trained on Within-Track Transitions.” arXiv, 2016. doi: 10.48550/ARXIV.1606.02096.

	[4] T. Mikolov, K. Chen, G. Corrado, and J. Dean, “Efficient Estimation of Word Representations in Vector Space.” arXiv, 2013. doi: 10.48550/ARXIV.1301.3781. 
}
